Alternative Methods for Retaining Explicit and Finding Implicit Sharing in Embedded DSLs

### 2023-02-15 paper/paper.tex
improper its
more its fixes

In the toy example
\begin{minted}{haskell}
class Exp repr where
  variable :: String -> repr Int
  constant :: String -> repr Int
  add :: repr Int -> repr Int -> repr Int
  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
the function \mintinline{haskell}{novel} will exhibit an issue in sharing of
it's two outputs when writing it's implementation that cannot be solved using
the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
issue in Section \ref{limitexplicit}. In our work, this translated into the
inability to process large library functions.

@@

In order to generate such a data structure, we will need to keep track of the
current maximum identifier to keep them unique. The representation for the
finally tagless instance is then a wrapper around a state monad that holds the
DAG being constructed in it's state and returns the current (top)
\mintinline{haskell}{NodeID}:

@@

The technique is essentially that of hash-consing, popularized by it's use in
LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}.
Other works have explored the use of type-safe hash-consing in embedded DSLs,
see \cite{filliatre:typesafeconsing}.

@@

The heart of our problem is that whenever we need to sequence the state of the inputs
for one of our DSL functions we want to first check if it's already been
evaluated. But how do we do that without first evaluating it to gain access to
it's unique identifier? We need some way to uniquely identify it outside the monad.

@@

To prevent confusion, we name the hash-consing function in our method \mintinline{haskell}{triecons}:
\begin{minted}{haskell}
triecons :: ByteString -> Node -> State DAG NodeID
triecons sAST node = do
  DAG trie maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
                   trie' = Trie.insert sAST (node,maxID') trie
                in do put $ DAG trie' maxID'
                      return maxID'
    Just (_,nodeID) -> return nodeID
\end{minted}
We use it to implement the DAG-building instance of the DSL,
which looks a lot like the previous instance.
The substantial differences are the 
\mintinline{haskell}{buildStringAST} calls
which you can think of as pretty printing, but optimized for the trie,
and the use of \mintinline{haskell}{seqArgs} (explained below):
\begin{minted}{haskell}
instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST "nadd" [e1,e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> triecons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST
\end{minted}
The magic is in \mintinline{haskell}{seqArgs}.
We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to find it's corresponding serialized AST in the Trie.
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}

@@

Even with explicit sharing via substitutions, our method contains a reasonable
amount of overhead in order to overcome the shortcomings of Kiselyov's method.
The \mintinline{haskell}{addChains} example altered for explicit sharing with
both methods presents a worst case scenario in terms of overhead comparison.
Kiselyov's method is able to fully utilize its' explicit sharing and our method
requires many substitution lookups.

@@

The method has it's drawbacks in terms of memory usage, but these can be
mitigated by explicitly specifying sharing. This does present an extra burden on
the DSL writer to implement explicit sharing when necessary and ensure labels
are not reused. Future work may investigate the use of a preprocessor or plugin
to automate explicit sharing. Unlike an explicit let construct, it would be
fairly straightforward to automatically bind \mintinline{haskell}{subT}
operations to any DSL function call.

>>>>

In the toy example
\begin{minted}{haskell}
class Exp repr where
  variable :: String -> repr Int
  constant :: String -> repr Int
  add :: repr Int -> repr Int -> repr Int
  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
the function \mintinline{haskell}{novel} will exhibit an issue in sharing of
its two outputs when writing its implementation that cannot be solved using
the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
issue in Section \ref{limitexplicit}. In our work, this translated into the
inability to process large library functions.

@@

In order to generate such a data structure, we will need to keep track of the
current maximum identifier to keep them unique. The representation for the
finally tagless instance is then a wrapper around a state monad that holds the
DAG being constructed in its state and returns the current (top)
\mintinline{haskell}{NodeID}:

@@

The technique is essentially that of hash-consing, popularized by its use in
LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}.
Other works have explored the use of type-safe hash-consing in embedded DSLs,
see \cite{filliatre:typesafeconsing}.

@@

The heart of our problem is that whenever we need to sequence the state of the inputs
for one of our DSL functions we want to first check if it's already been
evaluated. But how do we do that without first evaluating it to gain access to
its unique identifier? We need some way to uniquely identify it outside the monad.

@@

To prevent confusion, we name the hash-consing function in our method \mintinline{haskell}{triecons}:
\begin{minted}{haskell}
triecons :: ByteString -> Node -> State DAG NodeID
triecons sAST node = do
  DAG trie maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
                   trie' = Trie.insert sAST (node,maxID') trie
                in do put $ DAG trie' maxID'
                      return maxID'
    Just (_,nodeID) -> return nodeID
\end{minted}
We use it to implement the DAG-building instance of the DSL,
which looks a lot like the previous instance.
The substantial differences are the 
\mintinline{haskell}{buildStringAST} calls
which you can think of as pretty printing, but optimized for the trie,
and the use of \mintinline{haskell}{seqArgs} (explained below):
\begin{minted}{haskell}
instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST "nadd" [e1,e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> triecons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST
\end{minted}
The magic is in \mintinline{haskell}{seqArgs}.
We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to find its corresponding serialized AST in the Trie.
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}

@@

Even with explicit sharing via substitutions, our method contains a reasonable
amount of overhead in order to overcome the shortcomings of Kiselyov's method.
The \mintinline{haskell}{addChains} example altered for explicit sharing with
both methods presents a worst case scenario in terms of overhead comparison.
Kiselyov's method is able to fully utilize its explicit sharing and our method
requires many substitution lookups.

@@

The method has its drawbacks in terms of memory usage, but these can be
mitigated by explicitly specifying sharing. This does present an extra burden on
the DSL writer to implement explicit sharing when necessary and ensure labels
are not reused. Future work may investigate the use of a preprocessor or plugin
to automate explicit sharing. Unlike an explicit let construct, it would be
fairly straightforward to automatically bind \mintinline{haskell}{subT}
operations to any DSL function call.
